---
title: "EDA Lifelines Food Environment"
author: "Yamila Timmer"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    fig_caption: false
    lightbox: true
    gallery: true
    toc_depth: 4
    embed_fonts: true
    use_bookdown: true
    mathjax: "rmdformats"
    highlight: "tango"
    default_style: "light"
    downcute_theme: "default"
  
---

```{r, warning = FALSE, message = FALSE}
# Load libraries
library(tidyverse)
library(ggplot2)
library(dplyr)
library(stringr)
library(ggforce)
library(heatmaply)
library(UpSetR)
library(naniar)
```

After each looking into different variables that have to do with food environment, we have decided to continue with the following variables:

- Zipcode
- Body weight & body height (BMI)
- All nova food intakes (1-4)
- Added sugar intake
- Alcohol intake
- Kcal intake
- Presence of diabetes

Using these variables as a basis, we hope to be able to create machine learning algorithms to predict food intakes based on location and hope to further be able to link this to disease prevalence, such as diabetes or obesity. We came up with using the zipcode as an underlying factor, as we saw a consistent pattern on the lifelines interactive map, where certain health problems seemed more prevalent on the east side of the northern provinces, bordering Germany.

As we would like to make a user-friendly interactive data dashboard we first have to convert zipcodes to corresponding town- or city names. We are more interested in patterns within a whole town/city than specifically looking at differences between zip codes (as there might be very few participants per zipcode, not giving a realistic or reliable representation).

```{r}
# dataset for converting zipcode to town/city
zipcode_data <- read.csv("data/4pp.csv", header = TRUE)

# Lifelines dataset
data <- read.csv("data/Lifelines Dataset.csv ", header = TRUE, na.strings=c("NA","NaN", "")) # Convert NA, NaN and "" (blank field) to NA for uniformity
```


```{r}
# Selecting columns of interest
data_of_interest <- data %>%
  select(age_1a_q_1,
         
         gender_T1_M,
         zip_code_T1_M,
         bodylength_cm_all_m_1_T1_M,
         
         bodyweight_kg_all_m_1_T1_M,
         bodyweight_kg_all_m_1_T2_M,
         bodyweight_kg_all_m_1_T3_M,
         
         circumference_hip_all_m_1_T1_M,
         circumference_hip_all_m_1_T2_M,
         circumference_hip_all_m_1_T3_M,

         circumference_waist_all_m_1_T1_M,
         circumference_waist_all_m_1_T2_M,
         circumference_waist_all_m_1_T3_M,
         
         addedsugar_intake_adu_c_1_T1_QF,
         alcohol_intake_adu_c_1_T1_QF,
         kcal_intake_adu_c_1_T1_QF,
         nova_foodintake_adu_c_1_T1_QU,
         nova_foodintake_adu_c_4_T1_QU,
         
         diabetes_presence_adu_q_1_T1_Q1,
         diabetes_followup_adu_q_1_T2_Q1,
         diabetes_followup_adu_q_1_T3_Q1,
         
         cholesterol_result_all_m_1_T1_S,
         cholesterol_result_all_m_1_T2_S,
         cholesterol_result_all_m_1_T3_S,
         
         glucose_result_all_m_1_T3_S,
         glucose_result_all_m_1_T3_S,
         glucose_result_all_m_1_T3_S,
         
         bpavg_diastolic_all_m_1_T1_BP,
         bpavg_systolic_all_m_1_T1_BP,
         bpavg_diastolic_all_m_1_T2_BP,
         bpavg_systolic_all_m_1_T2_BP,
         bpavg_diastolic_all_m_1_T3_BP,
         bpavg_systolic_all_m_1_T3_BP
         )

```

As the names of the columns are quite long and thus a bit inconvenient to work with, we will rename the columns.
```{r}
data_of_interest <- data_of_interest %>%
  rename(zipcode = zip_code_T1_M,
         age = age_1a_q_1,
         gender = gender_T1_M,
         
         body_weight_1 = bodyweight_kg_all_m_1_T1_M,
         body_weight_2 = bodyweight_kg_all_m_1_T2_M,
         body_weight_3 = bodyweight_kg_all_m_1_T3_M,
         body_length_1 = bodylength_cm_all_m_1_T1_M,
         
         circumference_hip_1 = circumference_hip_all_m_1_T1_M,
         circumference_hip_2 = circumference_hip_all_m_1_T2_M,
         circumference_hip_3 = circumference_hip_all_m_1_T3_M,

         circumference_waist_1 = circumference_waist_all_m_1_T1_M,
         circumference_waist_2 = circumference_waist_all_m_1_T2_M,
         circumference_waist_3 = circumference_waist_all_m_1_T3_M,
         
         added_sugar = addedsugar_intake_adu_c_1_T1_QF,
         alcohol_intake = alcohol_intake_adu_c_1_T1_QF,
         kcal_intake = kcal_intake_adu_c_1_T1_QF,
         nova_foodintake_1 = nova_foodintake_adu_c_1_T1_QU,
         nova_foodintake_4 = nova_foodintake_adu_c_4_T1_QU,
         
         diabetes_presence_1 = diabetes_presence_adu_q_1_T1_Q1,
         diabetes_presence_2 = diabetes_followup_adu_q_1_T2_Q1,
         diabetes_presence_3 = diabetes_followup_adu_q_1_T3_Q1,
         
         cholesterol_1 = cholesterol_result_all_m_1_T1_S,
         cholesterol_2 = cholesterol_result_all_m_1_T2_S,
         cholesterol_3 = cholesterol_result_all_m_1_T3_S,
         
         glucose_1 = glucose_result_all_m_1_T3_S,
         glucose_2 = glucose_result_all_m_1_T3_S,
         glucose_3 = glucose_result_all_m_1_T3_S,
         
         diastolic_pressure_1 = bpavg_diastolic_all_m_1_T1_BP,
         systolic_pressure_1 = bpavg_systolic_all_m_1_T1_BP,
         diastolic_pressure_2 = bpavg_diastolic_all_m_1_T2_BP,
         systolic_pressure_2 = bpavg_systolic_all_m_1_T2_BP,
         diastolic_pressure_3 = bpavg_diastolic_all_m_1_T3_BP,
         systolic_pressure_3 = bpavg_systolic_all_m_1_T3_BP
         )
data_of_interest$age <- data_of_interest$age/12

```

Let's check the missing values, some zipcodes seem to be missing (`r sum(is.na(data_of_interest$zipcode))` to be precise), we will remove those before moving on. Other noteworthy commonly missing values are all measurements at T3, here ~50000 datapoints seem to be missing, which is quite a significant amount for the total of (`r nrow(data)`) rows. For T2 this seems to be ~20000 which is still significant, but less severe as T3. These missing values are caused by participants not coming back for assessments at T2 and T3, or by participants sadly passing away in the meantime. Depending on the machine learning model being trained, missing values can be dealt with by removing or imputing them. For now we will just make sure there are no missing values for creating the initial data set we will continue with, so for now it's important there are no missing zipcodes as we want to link the dataset to latitude/longitude data.
```{r}
gg_miss_var(data_of_interest) + 
  ylab("N. of missing values")
```
```{r}
data_of_interest <- data_of_interest %>%
  filter(!is.na(zipcode))

```


Below we join the "woonplaats" column of the zipcode dataset with the lifelines dataset, so that we have data on what zipcode corresponds to what city.
```{r}
data_of_interest <- data_of_interest %>%
    left_join(zipcode_data %>% select(latitude, longitude, postcode, woonplaats, gemeente, provincie), 
              by = c("zipcode" = "postcode"))

```

After merging the data with the cities, 3 zipcodes seem to have NA for city, meaning the zipcode was not present in the zipcode dataset, so we will manually have to add these following:

```
52.77472 6.49981 7928 Stuifzand Midden-Drenthe Drenthe
53.22564 6.65860 9613 Meerstad Groningen Groningen
52.74963 6.58211 7910 Nieuweroord Midden-Drenthe Drenthe
```

There also seem to be participants included from different regions, we remove all except for those from the provinces Drenthe, Groningen, Friesland.

```{r}
data_of_interest <- data_of_interest %>%
  filter(provincie %in% c("Drenthe", "Friesland", "Groningen"))
```

For diabetes presence, `1` is used to indicate diabetes and `2` is used to indicate that the patient does not have diabetes, for this scenario it would be more logical to use `0` to indicate absence of diabetes, so that is what we change below:
```{r}
data_of_interest$diabetes_presence_1[data_of_interest$diabetes_presence_1 == 2] <- 0
data_of_interest$diabetes_presence_2[data_of_interest$diabetes_presence_2 == 2] <- 0
data_of_interest$diabetes_presence_3[data_of_interest$diabetes_presence_3 == 2] <- 0

```

```{r}
data_of_interest$gender[data_of_interest$gender == "MALE"] <- 0
data_of_interest$gender[data_of_interest$gender == "FEMALE"] <- 1

```