---
title: "EDA NOVA and total kcal intake"
output: html_document
date: "2025-11-01"
---

```{r}
#libraries
library(tidyverse)
library(ggplot2)
library(dplyr)
library(stringr)
library(ggforce)
library(heatmaply)
library(UpSetR)
library(naniar)
```


```{r}
#processed data set with converted zipcodes
data_of_interest <- read.csv("data/data_of_interest.csv")


#Split data into east and west
threshold <- median(data_of_interest$longitude, na.rm = TRUE)
data_of_interest$region <- ifelse(data_of_interest$longitude > threshold, "East", "West")
```

Removing low kcal consumption
```{r}
#mean kcal intake
mean_kcal <- mean(data_of_interest$kcal_intake, na.rm = TRUE)
table_mean <- data.frame(Mean_kcal_intake = mean_kcal)
table_mean

data_of_interest %>%
  group_by(gender) %>%
  summarise(mean_kcal = mean(kcal_intake, na.rm = TRUE))

#cutoff low kcal intake (<500)
df_filtered <- data_of_interest %>%
  filter(kcal_intake >= 500)

new_mean_kcal <- mean(df_filtered$kcal_intake, na.rm = TRUE)
table_mean_new <- data.frame(new_mean_kcal_intake = new_mean_kcal)
table_mean_new

df_filtered %>%
  group_by(gender) %>%
  summarise(mean_kcal = mean(kcal_intake, na.rm = TRUE))
```


What is the difference in  calorie intake and (un)processed food consumption between the east and west?
```{r}
#Mean of calories and (un)processed food consumption.
data_of_interest %>%
  group_by(region) %>%
  summarise(
    mean_kcal = mean(kcal_intake, na.rm = TRUE),
            sd_kcal = sd(kcal_intake, na.rm = TRUE),
    
    mean_nova1 = mean(nova_foodintake_1, na.rm = TRUE),
    sd_nova1 = sd(nova_foodintake_1, na.rm = TRUE),
    
    mean_nova4 = mean(nova_foodintake_4, na.rm = TRUE),
    sd_nova4 = sd(nova_foodintake_4, na.rm = TRUE),
    
    count  = n()
  )

# Summarize means
summary_data <- data_of_interest %>%
  group_by(region) %>%
  summarise(
    mean_kcal = mean(kcal_intake, na.rm = TRUE),
    mean_nova1 = mean(nova_foodintake_1, na.rm = TRUE),
    mean_nova4 = mean(nova_foodintake_4, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = starts_with("mean_"), names_to = "variable", values_to = "value")

# Clean variable names
summary_data$variable <- gsub("mean_", "", summary_data$variable)

# Plot
ggplot(summary_data, aes(x = variable, y = value, fill = region)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of intake total kcal, unprocessed, and ultra-processed by east and west",
       x = "", y = "Intake (different units)") +
  theme_classic()
```
The east has higher total calorie intake, and consumes more processed and unprocessed food.

Is there a significant differences for calorie intake and (un)processed food consumption?
```{r}
t.test(kcal_intake ~ region, data = data_of_interest)
t.test(nova_foodintake_1 ~ region, data = data_of_interest)
t.test(nova_foodintake_3 ~ region, data = data_of_interest)
t.test(nova_foodintake_4 ~ region, data = data_of_interest)
```
Yes there are significant differences for kcal intake, nova1 and nova4 consumption. Noteably nova1 showed the highest P-value, showing it's the least significant.

What if splitting east and west by median is not the best, does longitude have effect on kcal intake and nova class?
```{r}
ggplot(data_of_interest, aes(x = longitude, y = kcal_intake)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(title = "Relationship between Longitude and kcal intake",
       x = "Longitude", y = "kcal intake") +
  theme_minimal()

ggplot(data_of_interest, aes(x = longitude, y = nova_foodintake_1)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(title = "Relationship between Longitude and kcal intake",
       x = "Longitude", y = "Unprocessed food intake (g/day)") +
  theme_minimal()

ggplot(data_of_interest, aes(x = longitude, y = nova_foodintake_4)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(title = "Relationship between Longitude and kcal intake",
       x = "Longitude", y = "Ultra-processed food intake (g/day)") +
  theme_minimal()
```

```{r}
# Reshape data into long format
data_long <- data_of_interest %>%
  pivot_longer(
    cols = c(kcal_intake, nova_foodintake_1, nova_foodintake_4),
    names_to = "variable",
    values_to = "intake"
  )

# Define readable labels
variable_labels <- c(
  kcal_intake = "Kcal intake (kcal/day)",
  nova_foodintake_1 = "Unprocessed food (g/day)",
  nova_foodintake_4 = "Ultra-processed food (g/day)"
)

# Plot trendlines
ggplot(data_long, aes(x = longitude, y = intake, color = variable)) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2) +
  scale_color_manual(
    values = c(
      "kcal_intake" = "#E41A1C",      # red
      "nova_foodintake_1" = "#377EB8", # blue
      "nova_foodintake_4" = "#4DAF4A"  # green
    ),
    labels = variable_labels
  ) +
  labs(
    title = "Food intake vs Longitude",
    x = "Longitude",
    y = "Intake (different units)",
    color = "Variable"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )


```

