---
title: "Lifelines Food Environment: Alcohol Intake"
author: "Yamila Timmer"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    fig_caption: false
    lightbox: true
    gallery: true
    toc_depth: 4
    embed_fonts: true
    use_bookdown: true
    mathjax: "rmdformats"
    highlight: "tango"
    default_style: "light"
    downcute_theme: "default"
---

```{r}
library(plotly)
library(ggplot2)

```

# Visualizing

Here, I will generate some visualisations to look at alcohol intake in the lifelines dataset, specifically I will be looking at 'total alcohol intake in g/day'.

For comparison:

- 1 standard glass of beer (250mL) contains 10 gram of alcohol


```{r}
summary(data_of_interest$alcohol_intake)
```


## Gender vs Alcohol

```{r}
ggplot(data_of_interest, aes(x = gender, y = alcohol_intake, fill = gender)) + 
  geom_boxplot() +
  xlab("Gender") +
  ylab("Daily alcohol intake in gram")

```
Conclusions:

- There seem to be many outliers, for both men and women, so it seems wise to leave these in as they don't seem to be an exception
- Alcohol consumption seems to be higher among men


```{r}
# Visualize age distribution/gender
ggplot(data = data_of_interest) + 
  geom_density(mapping = aes(x = alcohol_intake, 
                             color = gender, 
                             fill = gender), alpha = 0.2) + 
  xlab("Daily alcohol intake in gram") + 
  ggtitle("Distribution of alcohol intake per gender")
```
Conclusions:

- There seem to be a lot more women that dont drink on a daily basis, compared to men
- Around 35-40 seems to be a cutoff point, above this point there are very little values

## Age vs Alcohol

```{r}
ggplot(data_of_interest, aes(x=age, y=alcohol_intake)) + 
    geom_point() +
  xlab("Age in years") + 
  ylab("Daily alcohol intake in gram") +
  ggtitle("Distribution of alcohol intake across age")

```
Conclusions:

- It seems like alcohol intake does not differ that much across age, except for participants above 75 years, there seem to be a lot less high values here

## Alcohol vs City

```{r}
city_data <- data_of_interest %>%
  group_by(woonplaats) %>%
  summarise(avg_alcohol = mean(alcohol_intake))

```

```{r}

city_alcohol <- data_of_interest %>%
  group_by(woonplaats) %>%
  summarise(mean_alcohol = mean(alcohol_intake, na.rm = TRUE)) %>%
  arrange(desc(mean_alcohol)) %>%
  slice_head(n = 10)


ggplot(city_alcohol, aes(x = reorder(woonplaats, mean_alcohol), y = 1, fill = mean_alcohol)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "inferno", name = "Avg. alcohol intake") +
  labs(title = "Top 10 Cities by Alcohol Intake",
       x = "City", y = "") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank()
  )

```
```{r}
ggplot(data_of_interest, aes(x = provincie, y = alcohol_intake, fill = provincie)) +
  geom_violin(trim = FALSE) +
  labs(
    title = "Distribution of Alcohol Intake per Province",
    x = "Province",
    y = "Alcohol Intake"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "none"  # remove redundant color legend
  )


```

```{r}
data <- as.matrix(data_of_interest)
heatmap(data_of_interest)

```

# Map


```{r}
library(leaflet)
library(mapview)

```

Divide avg intake by participants, to normalize?
```{r}
city_data <- data_of_interest %>%
  group_by(woonplaats) %>%
  summarise(avg_alcohol = mean(alcohol_intake))

```

```{r}
city_data <- city_data %>%
    left_join(zipcode_data %>% select(latitude, longitude, woonplaats), 
              by = c("woonplaats" = "woonplaats"))

```

```{r}
m = leaflet(data = city_data) %>%
   addTiles() %>%
   addCircleMarkers(~longitude, ~latitude , popup = ~as.character(avg_alcohol))
m

```


```{r}
mapview(data_of_interest, xcol = "longitude", ycol = "latitude", crs = 4269, grid = FALSE)
```
