---
title: "Lifelines Food Environment: Predicting Disease"
author: "Yamila Timmer"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    fig_caption: false
    lightbox: true
    gallery: true
    toc_depth: 4
    embed_fonts: true
    use_bookdown: true
    mathjax: "rmdformats"
    highlight: "tango"
    default_style: "light"
    downcute_theme: "default"
---

In this notebook I will make a categorical machine learning model that can predict disease risk. We are using the Lifelines data set. From the available data there are two diseases that can be related to dietary choices, which are diabetes and hypertension.

The data of interest has already been constructed after running the data pre-processing R markdown file, however to train specific models for diabetes/hypertension prediction, some missing values will have to be dealt with.

# Diabetes prediction

The dataset contains 3 variables needed to determine diabetes occurence, which are the 3 seperate diabetes 'checks'. Lifelines has 3 different timepoints, each around 5-10 years apart. For the first timepoint the participant is asked whether they have diabetes. Here 2 equals 'no' and 1 equals 'yes'. Then at T2, several years later, participants are asked if they have developed diabetes in the time between T1 and T2. Again here 2 equals 'no' and 1 equals 'yes'. Lastly, the participants are asked at T3 if they have developed diabetes between T2 and T3.
```{r}
data_of_interest <- data_of_interest %>%
  filter(!is.na(gender)) %>%
  filter(!is.na(diabetes_presence_2)) %>%
  filter(!is.na(diabetes_presence_1)) %>%
  filter(!is.na(diabetes_presence_3)) %>%
  filter(!is.na(nova_foodintake_4))

```

```{r}
library(tidyverse)

data_long <- data_of_interest %>%
  pivot_longer(
    cols = starts_with("diabetes_presence"),
    names_to = "variable",
    values_to = "value"
  ) %>%
  filter(value == 1)

ggplot(data_long, aes(x = variable)) +
  geom_bar() +
  theme_minimal() +
  labs(
    x = "Timepoints",
    y = "Amount of participants",
    title = "Amount of new cases of diabetes per timepoint"
  )

```

To be able to predict diabetes occurrence, we need a decent amount of data on participants that developed diabetes during Lifelines. We do not look at T1, as the survey made no distinction between diabetes type I and type II. As only type II is heavily correlated with food environment, we are only interested in this type. Type II often develops later in life, thus we will look only at T2 and T3. To get a more complete overview of what and how many participants develop diabetes during the complete duration of Lifelines we combine diabetes_presence_2 and diabetes_presence_3.

```{r}
data_of_interest <- data_of_interest %>%
  mutate(
    new_diabetes = if_else(
      diabetes_presence_2 == 1 |
      diabetes_presence_3 == 1,
      1, 
      2 
    )
  )

```


```{r}
gg_miss_var(data_of_interest) + 
  ylab("N. of missing values")

```


# Machine learning model

```{r}
# -------------------------------------------------
# 0. Load packages
# -------------------------------------------------
library(tidyverse)
library(caret)
library(smotefamily)
library(randomForest)
library(pROC)
library(PRROC)
```

```{r}
# -------------------------------------------------
# 1. Load and prepare data
# -------------------------------------------------
machine_learning_data <- data_of_interest

# Choose predictor variables
predictor_vars <- c(
  "latitude", "longitude", "age", "body_weight_1", 
  "circumference_hip_1", "circumference_waist_1", 
  "kcal_intake", "nova_foodintake_1", "nova_foodintake_4"
)

machine_learning_data <- machine_learning_data %>%
  select(all_of(c(predictor_vars, "new_diabetes")))

# -------------------------------------------------
# 2. Train/test split
# -------------------------------------------------
set.seed(123)
train_index <- createDataPartition(machine_learning_data$new_diabetes, p = 0.8, list = FALSE)
train_data <- machine_learning_data[train_index, ]
test_data  <- machine_learning_data[-train_index, ]

# Convert outcome to factor
train_data$new_diabetes <- factor(ifelse(train_data$new_diabetes == 1, "Yes", "No"), levels = c("No","Yes"))
test_data$new_diabetes  <- factor(ifelse(test_data$new_diabetes == 1, "Yes", "No"), levels = c("No","Yes"))

# -------------------------------------------------
# 3. Dummy encode predictors
# -------------------------------------------------
dummies <- dummyVars(~ ., data = train_data %>% select(-new_diabetes), fullRank = FALSE)
X_num <- predict(dummies, newdata = train_data %>% select(-new_diabetes)) %>% as.data.frame()
y <- train_data$new_diabetes

# -------------------------------------------------
# 4. Apply SMOTE for class balancing
# -------------------------------------------------
balanced <- SMOTE(X_num, y, K = 5)
train_balanced <- balanced$data
colnames(train_balanced)[ncol(train_balanced)] <- "new_diabetes"
train_balanced$new_diabetes <- factor(train_balanced$new_diabetes, levels = c("No","Yes"))

# -------------------------------------------------
# 5. Train RandomForest with higher weight for 'Yes'
# -------------------------------------------------
library(randomForest)
model <- randomForest(
  new_diabetes ~ .,
  data = train_balanced,
  classwt = c("No" = 1, "Yes" = 100)  # increases sensitivity for Yes
)

# -------------------------------------------------
# 6. Prepare test set
# -------------------------------------------------
test_num <- predict(dummies, newdata = test_data %>% select(-new_diabetes)) %>% as.data.frame()
test_num$new_diabetes <- test_data$new_diabetes

# -------------------------------------------------
# 7. Predict probabilities
# -------------------------------------------------
probs <- predict(model, newdata = test_num, type = "prob")
prob_yes <- probs[, "Yes"]  # extract probability for 'Yes'

# -------------------------------------------------
# 8. ROC curve
# -------------------------------------------------
library(pROC)
roc_obj <- roc(test_num$new_diabetes, prob_yes, levels = c("No","Yes"))
plot(roc_obj, col = "blue", lwd = 2, main = "ROC Curve")
cat("AUC:", auc(roc_obj), "\n")

# -------------------------------------------------
# 9. Choose threshold for high sensitivity
# -------------------------------------------------
coords_sens <- coords(roc_obj, x = "all", ret = c("threshold", "sensitivity", "specificity"))
high_sens_threshold <- 0.05
cat("Threshold for high sensitivity:", high_sens_threshold, "\n")

# -------------------------------------------------
# 10. Final evaluation using chosen threshold
# -------------------------------------------------
custom_pred <- ifelse(prob_yes > high_sens_threshold, "Yes", "No")
confusionMatrix(factor(custom_pred, levels=c("No","Yes")), test_num$new_diabetes, positive="Yes")

# -------------------------------------------------
# 11. Optional: F1 score threshold optimization
# -------------------------------------------------
thresholds <- seq(0, 1, by = 0.01)
f1_scores <- sapply(thresholds, function(t) {
  pred <- ifelse(prob_yes > t, "Yes", "No")
  cm <- confusionMatrix(factor(pred, levels=c("No","Yes")), test_num$new_diabetes, positive="Yes")
  precision <- cm$byClass["Pos Pred Value"]
  recall <- cm$byClass["Sensitivity"]
  if (is.na(precision) | is.na(recall) | (precision + recall) == 0) return(0)
  2 * (precision * recall) / (precision + recall)
})
best_f1_threshold <- thresholds[which.max(f1_scores)]
cat("Best threshold (F1 score):", best_f1_threshold, "\n")

custom_pred_f1 <- ifelse(prob_yes > best_f1_threshold, "Yes", "No")
confusionMatrix(factor(custom_pred_f1, levels=c("No","Yes")), test_num$new_diabetes, positive="Yes")
```

```{r}
new_person <- data.frame(
  latitude = 53.2,
  longitude = 6.6,
  age = 45,
  body_weight_1 = 158,
  circumference_waist_1 = 142,
  circumference_hip_1 = 110,
  kcal_intake = 4400,
  nova_foodintake_1 = 100,
  nova_foodintake_4 = 300
  
)


new_person_num <- predict(dummies, newdata = new_person) %>% as.data.frame()


# Make sure missing dummy columns are added with 0
missing_cols <- setdiff(names(train_balanced)[-ncol(train_balanced)], names(new_person_num))
for (col in missing_cols) new_person_num[[col]] <- 0

# Predict
pred_class <- predict(model, newdata = new_person_num)
pred_prob  <- predict(model, newdata = new_person_num, type = "prob")

print(pred_class)
print(pred_prob)

```

TODO:
- It is difficult to predict true as true, because there are a lot more negative cases, maybe ask Fenna for a way to improve
- join with T_3 diabetes
- Add new parameters


# Hypertension
## Background info

Another disease, that can also be related to food environment, that we can add is hypertension. According to the national heart, lung and blood institute of the United States, a person has hypertension if their blood pressure is 130/80 or higher. With the first number being the systolic pressure and the second the diastolic pressure. We have data on both so we can look into how many people suffer from hypertension.

```{r}
# -------------------------------
# 1. Define hypertension outcome
# -------------------------------
# Hypertension: systolic ≥130 OR diastolic ≥80
data_of_interest <- data_of_interest %>%
  mutate(
    hypertension_t1 = if_else(systolic_pressure_1 >= 130 | diastolic_pressure_1 >= 80, 1, 0),
    hypertension_t2 = if_else(systolic_pressure_2 >= 130 | diastolic_pressure_2 >= 80, 1, 0),
    hypertension_t3 = if_else(systolic_pressure_3 >= 130 | diastolic_pressure_3 >= 80, 1, 0),
    hypertension = case_when(
      is.na(hypertension_t2) & is.na(hypertension_t3) ~ hypertension_t1,          # use T1 if T2 & T3 NA
      is.na(hypertension_t3) & !is.na(hypertension_t2) ~ hypertension_t2,         # use T2 if T3 NA
      TRUE ~ if_else(hypertension_t1 == 1 | hypertension_t2 == 1 | hypertension_t3 == 1, 1, 0)  # otherwise 1 if any = 1
    )
  )


```

```{r}

# -------------------------------
# 2. Prepare predictor variables
# -------------------------------
predictor_vars <- c(
  "latitude", "longitude", "age", "body_weight_1",
  "circumference_hip_1", "circumference_waist_1",
  "kcal_intake", "nova_foodintake_1", "nova_foodintake_4"
)

ml_data <- data_of_interest %>%
  select(all_of(c(predictor_vars, "hypertension")))

# -------------------------------
# 3. Train/Test Split
# -------------------------------
set.seed(123)
train_index <- createDataPartition(ml_data$hypertension, p = 0.8, list = FALSE)
train_data <- ml_data[train_index, ]
test_data  <- ml_data[-train_index, ]

# Convert outcome to factor
train_data$hypertension <- factor(ifelse(train_data$hypertension == 1, "Yes", "No"), levels=c("No","Yes"))
test_data$hypertension  <- factor(ifelse(test_data$hypertension == 1, "Yes", "No"), levels=c("No","Yes"))

# -------------------------------
# 4. Dummy encode predictors
# -------------------------------
dummies <- dummyVars(~ ., data = train_data %>% select(-hypertension), fullRank = FALSE)
X_num <- predict(dummies, newdata = train_data %>% select(-hypertension)) %>% as.data.frame()
y <- train_data$hypertension

# -------------------------------
# 5. Apply SMOTE for class balancing
# -------------------------------
balanced <- SMOTE(X_num, y, K = 5)
train_balanced <- balanced$data
colnames(train_balanced)[ncol(train_balanced)] <- "hypertension"
train_balanced$hypertension <- factor(train_balanced$hypertension, levels=c("No","Yes"))

# -------------------------------
# 6. Train RandomForest with class weights
# -------------------------------
model <- randomForest(
  hypertension ~ .,
  data = train_balanced,
  classwt = c("No"=1, "Yes"=5)  # increase weight for Yes to improve sensitivity
)

# -------------------------------
# 7. Prepare test set
# -------------------------------
test_num <- predict(dummies, newdata = test_data %>% select(-hypertension)) %>% as.data.frame()
test_num$hypertension <- test_data$hypertension

# -------------------------------
# 8. Predict probabilities
# -------------------------------
probs <- predict(model, newdata = test_num, type = "prob")
prob_yes <- probs[, "Yes"]

# -------------------------------
# 9. ROC Curve and AUC
# -------------------------------
roc_obj <- roc(test_num$hypertension, prob_yes, levels=c("No","Yes"))
plot(roc_obj, col="blue", lwd=2, main="ROC Curve")
cat("AUC:", auc(roc_obj), "\n")

# -------------------------------
# 10. Select threshold for high sensitivity
# -------------------------------
coords_sens <- coords(roc_obj, x="all", ret=c("threshold", "sensitivity", "specificity"))
high_sens_threshold <- coords_sens$threshold[which.max(coords_sens$sensitivity)]
cat("Threshold for high sensitivity:", high_sens_threshold, "\n")

# -------------------------------
# 11. Evaluate model at chosen threshold
# -------------------------------
custom_pred <- ifelse(prob_yes > high_sens_threshold, "Yes", "No")
confusionMatrix(factor(custom_pred, levels=c("No","Yes")), test_num$hypertension, positive="Yes")
```

```{r}
# -------------------------------
# 12. Predict new individual
# -------------------------------
new_person <- data.frame(
  latitude = 53.2,
  longitude = 6.6,
  age = 80,
  body_weight_1 = 200,
  circumference_waist_1 = 150,
  circumference_hip_1 = 150,
  kcal_intake = 2500,
  nova_foodintake_1 = 100,
  nova_foodintake_4 = 3000
)

new_person_num <- predict(dummies, newdata = new_person) %>% as.data.frame()

# Add missing dummy columns
missing_cols <- setdiff(names(train_balanced)[-ncol(train_balanced)], names(new_person_num))
for (col in missing_cols) new_person_num[[col]] <- 0

# Predict probability for new person
prob_new <- predict(model, newdata = new_person_num, type="prob")[, "Yes"]
cat("Predicted chance of developing hypertension in next 10 years:", round(prob_new*100,2), "%\n")



```


## Distribution of hypertension
