---
title: "Lifelines Food Environment: Predicting Disease"
author: "Yamila Timmer"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    fig_caption: false
    lightbox: true
    gallery: true
    toc_depth: 4
    embed_fonts: true
    use_bookdown: true
    mathjax: "rmdformats"
    highlight: "tango"
    default_style: "light"
    downcute_theme: "default"
---

In this notebook I will make a categorical machine learning model that can predict disease risk. We are using the Lifelines data set. From the available data there are two diseases that can be related to dietary choices, which are diabetes and hypertension.

The data of interest has already been constructed after running the data pre-processing R markdown file, however to train specific models for diabetes/hypertension prediction, some missing values will have to be dealt with.

```{r}
#imports
library(tidyverse)
library(tidyverse)
library(pROC)
```

# Diabetes prediction

The dataset contains 3 variables needed to determine diabetes occurence, which are the 3 seperate diabetes 'checks'. Lifelines has 3 different timepoints, each around 5-10 years apart. For the first timepoint the participant is asked whether they have diabetes. Here 2 equals 'no' and 1 equals 'yes'. Then at T2, several years later, participants are asked if they have developed diabetes in the time between T1 and T2. Again here 2 equals 'no' and 1 equals 'yes'. Lastly, the participants are asked at T3 if they have developed diabetes between T2 and T3. To be able to predict diabetes occurrence, we need a decent amount of data on participants that developed diabetes during Lifelines. We do not look at T1, as the survey made no distinction between diabetes type I and type II. As only type II is heavily correlated with food environment, we are only interested in this type. Type II often develops later in life. For this model we focus on patients that develop diabetes between T1 and T2. 

```{r}
gg_miss_var(data_of_interest) + 
  ylab("N. of missing values")

```

In total `r sum(data_of_interest$new_diabetes == 1)` people have diabetes at at least one point during Lifelines, meaning they could be a good predictor for whether other people are at risk of also developing diabetes.

## Machine learning model

```{r}
# =================================================
# 1. DATA PREPARATION
# =================================================

# Make new onset column for people that develop diabetes between T1/T2
diabetes_data <- data_of_interest %>%
  filter(
    !is.na(diabetes_presence_1),
    !is.na(diabetes_presence_2)
  ) %>%

  mutate(
    diabetes_new_onset = if_else(
      diabetes_presence_1 == 0 & diabetes_presence_2 == 1, 1, 0
    ),
  ) %>%
  filter(diabetes_presence_1 == 0) 
```


```{r}
# =================================================
# 2. LOGISTIC REGRESSION MODEL
# =================================================

# Create model and select predictor variables
model <- glm(
  diabetes_new_onset ~ 
    age +
    bmi_t1 +
    circumference_waist_1 + 
    gender + 
    circumference_hip_1 + 
    latitude + 
    longitude +
    alcohol_intake + 
    nova_foodintake_1 +   
    nova_foodintake_4,
  data = diabetes_data,
  family = binomial
)

summary(model)

```

```{r}
# =================================================
# 3. ROC + AUC
# =================================================

# Predicted probabilities for the dataset
pred_probs <- predict(model, newdata = diabetes_data, type = "response")

# ROC object
roc_obj <- roc(
  response = diabetes_data$diabetes_new_onset,
  predictor = pred_probs,
  quiet = TRUE
)

# Print AUC
print(paste("AUC:", auc(roc_obj)))

# Plot ROC curve
plot(
  roc_obj,
  col = "blue",
  lwd = 3,
  main = "ROC Curve for Logistic Regression Model"
)

```

```{r}
# =================================================
# 4. CONFUSION MATRIX
# =================================================

# Low threshold is chosen, as new onset diabetes is very rare 
threshold <- 0.03

# Convert probabilities to predicted classes
pred_class <- ifelse(pred_probs >= threshold, 1, 0)

# Create confusion matrix
conf_matrix <- table(
  Actual = diabetes_data$diabetes_new_onset,
  Predicted = pred_class
)

conf_matrix


```

Now we can try predicting new cases, below there are examples of a 'person' with low-risk and high-risk.

```{r}
# =================================================
# 5. PREDICT NEW PERSON
# =================================================

# Low risk (e.g. healthy body mass, young age, etc.)
low_risk_person <- data.frame(
  age = 22,
  gender = 0,                     # female
  circumference_waist_1 = 65,
  circumference_hip_1 = 90,
  added_sugar = 5,
  alcohol_intake = 0,
  latitude = 52,
  longitude = 6,
  nova_foodintake_1 = 12,         # high unprocessed food intake
  nova_foodintake_4 = 1,          # very low ultra-processed intake
  bmi_t1 = 20                     # healthy BMI
)


# High risk (e.g. high body mass, high age, etc.)
high_risk_person <- data.frame(
  age = 78,
  gender = 1,                     # male (higher risk in your model)
  circumference_waist_1 = 135,    # extreme abdominal obesity
  circumference_hip_1 = 105,
  added_sugar = 40,
  alcohol_intake = 20,
  latitude = 52,
  longitude = 8,
  nova_foodintake_1 = 0,          # almost no unprocessed food
  nova_foodintake_4 = 12,         # very high ultra-processed intake
  bmi_t1 = 42                     # severe obesity
)



prob_low_risk <- predict(model, newdata = low_risk_person, type = "response")
prob_high_risk <- predict(model, newdata = high_risk_person, type = "response")

print(paste("Predicted risk (low risk person):", round(prob_low_risk * 100, 2), "%"))
print(paste("Predicted risk (high risk person):", round(prob_high_risk * 100, 2), "%"))

```

As expected, the risk of developing diabetes is much lower for the low risk person and relatively high for the high risk person. This percentage shows the risk of developing diabetes in the next 5-10 years, as that is the approximate time for people between T1 and T2.



# Hypertension
## Background info

Another disease, that can also be related to food environment, that we can add is hypertension. According to the national heart, lung and blood institute of the United States, a person has hypertension if their blood pressure is 130/80 or higher. With the first number being the systolic pressure and the second the diastolic pressure. We have data on both so we can look into how many people suffer from hypertension.



## Machine learning model

```{r}
# =================================================
# 1. DATA PREPARATION
# =================================================

hypertension_data <- data_of_interest %>%
  filter(
    !is.na(systolic_pressure_2),
    !is.na(diastolic_pressure_2)
  ) %>%
  mutate(
    hypertension_t1 = if_else(
      systolic_pressure_1 >= 130 | diastolic_pressure_1 >= 80, 1, 0
    ),
    hypertension_t2 = if_else(
      systolic_pressure_2 >= 130 | diastolic_pressure_2 >= 80, 1, 0
    ),
    hypertension_new_onset = if_else(
      hypertension_t1 == 0 & hypertension_t2 == 1, 1, 0
    ),
  ) %>%
  filter(hypertension_t1 == 0)

```



```{r}
# =================================================
# 2. LOGISTIC REGRESSION MODEL
# =================================================

model <- glm(
  hypertension_new_onset ~ 
    age + 
    gender + 
    circumference_waist_1 + 
    latitude + 
    alcohol_intake + 
    nova_foodintake_1 +  
    bmi_t1,
  data = hypertension_data,
  family = binomial
)

summary(model)

```

```{r}
# =================================================
# 3. ROC + AUC
# =================================================

# Predicted probabilities for the dataset
pred_probs <- predict(model, newdata = hypertension_data, type = "response")

# ROC object
roc_obj <- roc(
  response = hypertension_data$hypertension_new_onset,
  predictor = pred_probs,
  quiet = TRUE
)

# Print AUC
print(paste("AUC:", auc(roc_obj)))

# Plot ROC curve
plot(
  roc_obj,
  col = "blue",
  lwd = 3,
  main = "ROC Curve for Logistic Regression Model"
)

```

```{r}
# =================================================
# 4. PREDICT NEW PERSON
# =================================================
new_person <- data.frame(
  age = 20,
  gender = 0,                     # female
  circumference_waist_1 = 65,
  circumference_hip_1 = 85,
  added_sugar = 5,
  alcohol_intake = 0,
  latitude = 53,                  # higher latitude → lower risk in your model
  longitude = 6,
  nova_foodintake_1 = 8,
  nova_foodintake_4 = 2,
  bmi_t1 = 30,
  diastolic_pressure_1 = 80,
  systolic_pressure_1 = 120
)


prob <- predict(model, newdata = new_person, type = "response")
print(paste("Predicted risk:", round(prob * 100, 2), "%"))


```





If we look at a summary of the hypertension model, we can see what factors seem the most significant for the outcome, so which factors have the most meaningful effect on whether someone develops hypertension or not. The most significant factors seem to include: age, gender, bmi, waist circumference and cholesterol. Nova food intake 1 (non-processed food intake) seems to be significant, but not as severe as the aforementioned factors. Caloric intake, diabetes presence, added sugar, alcohol intake, latitude, longitude, hip circumference and nova food intake 4 all do not seem significant (p > 0.05)



Looking at the model summary, we see the estimated size and direction of the effect, the standard error, the z-value (estimate/std error) and the p-value, which shows how likely the effect is caused by pure chance. Age has a positive estimate, indicating that risk increases with old age and seems to be a strong predictor, looking at the low p-value. Gender has a negative estimate, meaning the higher the number, the lower the risk. In this case, for age there are only two possible numbers, 0 for male and 1 for female. Meaning that the risk of hypertension is lower for women than for men. 


Write something on limitations
- Would have been better if there was more nutritional data to work with from more timepoints, as things like diet are highly subject to change, so only T1 and T2 were used, as the time between T1 and T3 is over 10 years
- Cholesterol was also a significant predictor (p < 0.05), but not easily measurable at home so not used for our application

clinical definition is >80 or >130 not and https://www.ncbi.nlm.nih.gov/books/NBK539859/

if person is already hypertensive, they should get a message/popup, current bloodpressure indicates hypertension. Also info on how to measure bloodpressure?

Else get percentage chance for how likely they are to develop hypertension in the next 5-10 years

